A protocol \& event-sourced database for decentralized user-siloed data

### 2019-09-29 draft.tex
more edits
Signed-off-by: andrew <andrew@textile.io>

While companies collecting user data may not itself be a significant problem, it has been shown that, over time company incentives often shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}. When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value/revenue for the company. It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data. Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data. This approach is called \emph{user-siloed} data and it fundamentally separates apps from their users' data.

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have addressed the problem directly, by providing a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.

@@

We now describe some of the technologies and concepts motivating the design of our novel decentralized database system. We highlight some of the advantages and lessons learned from event sourcing (ES) and discuss drawbacks to using these approaches in decentralized systems. We provide an overview of some important technologies related to the Interplanetary File System (IPFS) that make it possible to rethink ES in a decentralized network. Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are used in popular databases outside the IPFS context.

@@

When developing large-scale software systems, it is common to store data in a relational database management system (RDBMS). To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database. A powerful alternative is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order. This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state. This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{jaykrepsLogWhatEvery2013}.

@@

With these interface simplifications, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems. Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads. For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods. The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example). These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.

@@

Similar abstractions could (and will) be used to implement additional database types and functions. Tables, feeds, counters, and other simple stores can also be built on Threads. Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand. Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these ``example'' libraries.

@@

At an abstract level, the Textile protocol provides a distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable. Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls. Indeed, by designing our system around Events, a Dispatcher, and generic Stores, we make it easy to extend Threads in many different ways. Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.

@@

We show that the flexible core structure of single-writer append-only logs can be used to compose higher-order structures such as Threads, Views, and/or CRDTs. In particular, we show that through the design of specific default view Models, we can support important features such as access control lists and common and/or complex data models. The Threads protocol described here is flexible enough to derive numerous specific database types (e.g. key/value stores, document stores, relational stores, etc) and model an unlimited number of applications states. The cryptography used throughout Threads will help shift the data ownership model from apps to users.

@@

Threads change the relationship between a user, their data, and the services they connect with that data. The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, on a network of Textile peers. Based on the Follow Key now available in Threads, Bots would be able to relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less} way. Bots can additionally enhance the IPFS network by providing a framework to build and deploy many kinds of trust-less or trusted services. Examples include simple archival, caching and republishing Bots. Other examples include payment, re-encryption, or bridges to Web 2.0 services to offer decentralized access to Web 2.0.

@@

Implementing Threads internals (see Section \ref{sec:internals}) using similar patterns to common frontend workflows (e.g., Redux) presents opportunities for tighter coupling between ``backend'' logic and fontend views. This is a major advantage of tools such as reSolve\footnote{\url{https://reimagined.github.io/resolve/}}, where ``system changes can be reflected immediately [on the frontend], without the need to re-query the backend'' \cite{ereminReduxInspiredBackend2019}. Textile developers will develop frameworks to more directly expose the internals of Threads to frontend SDKs, making it possible to sync application state across and between apps and services on the Textile network.

>>>>

While the fact that companies collect user data may not itself be a significant problem, problems may arise if over time a company's incentives shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}. When this incentive shift happens, companies that have been creating data-silos may treat that data as new source of value or revenue for the company. It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data. Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data. This approach is called \emph{user-siloed} data and it fundamentally separates apps from their users' data.

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have addressed the problem of \emph{how data should be collected}, by providing a sufficiently interoperable protocol for scalable storage, transmission, and use of user-siloed application data.

@@

We now describe some of the technologies and concepts motivating the design of our novel decentralized database system. We highlight some of the advantages and lessons learned from event sourcing (ES) and discuss drawbacks to using these approaches in decentralized systems. We provide an overview of some important technologies related to the Interplanetary File System (IPFS) that make it possible to rethink ES in a decentralized network. Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are designed in popular database management (DBMS) systems outside the IPFS context.

@@

To model realistic systems, apps often need to map data between domain models and database tables, where the same data model is used to both query and update a database. To solve \emph{synchronization} it is often helpful to handle updates on just the database and then provide the query and interfaces only later in a DBMS. One powerful approach to synchronization is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order. This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state. This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{jaykrepsLogWhatEvery2013}.

@@

With these interface simplifications, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores or DBMS. Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads. For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods. The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example). These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.

@@

Similar abstractions could (and will) be used to implement additional database types and functions. Tables, feeds, counters, and other simple stores can also be built on Threads. Each database style would be implemented as a standalone wrapper/software library, allowing application developers to pick and choose the solution most useful to the application at hand. Similarly, more advanced applications could be implemented using a combination of database types, or by examining the source code of these \emph{reference} libraries.

@@

The Textile protocol provides a distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable. Any internal implementation details of a compliant Threads \emph{client} may use any number of well-established design patterns from the CQRS and ES (and related) literature to \emph{extend} the Threads protocol with additional features and controls. Indeed, by designing our system around Events, a Dispatcher, and generic Stores, we make it easy to extend Threads in many different ways. Some extensions included by Textile's Threads implementation are outlined in this section to provide some understanding of the extensibility this design affords.

@@

We show that the flexible core structure of single-writer append-only logs can be used to compose higher-order structures such as Threads, Views, and/or CRDTs. In particular, we show that through the design of specific default view Models, we can support important features such as access control lists and common, specialized, or complex data models. The Threads protocol described here is flexible enough to derive numerous specific database types (e.g. key/value stores, document stores, relational stores, etc) and model an unlimited number of applications states. The cryptography used throughout Threads will help shift the data ownership model from apps to users.

@@

@@

Implementing Threads internals (see Section \ref{sec:internals}) using similar patterns to common frontend workflows (e.g., Redux) presents opportunities for tighter coupling between ``backend'' logic and fontend views. This is a major advantage of tools such as reSolve\footnote{\url{https://reimagined.github.io/resolve/}}, where ``system changes can be reflected immediately [on the frontend], without the need to re-query the backend'' \cite{ereminReduxInspiredBackend2019}. Textile developers will create frameworks to more directly expose the internals of Threads to frontend SDKs (or DMBS, see \ref{appx:DBMS}), making it possible to sync application state across and between apps and services on the IPFS network.

Threads change the relationship between a user, their data, and the services they connect with that data. The nested, or multi-layered, encryption combined with powerful ACL capabilities create new opportunities to build distributed services, or Bots, in the network of IPFS peers. Based on the Follow Key now available in Threads, Bots can relay, replicate, or store data that is synchronized via real-time updates in a \emph{trust-less}, partially trusted, or fully-trusted way. Bots can additionally enhance the IPFS network by providing a framework to build and deploy many new kinds of services available over HTTP or P2P. Services could include simple data archival, caching and republishing, translation, data conversion, and more. Advanced examples could include payment, re-encryption, or bridges to Web 2.0 services to offer decentralized access to Web 2.0.
